---
title: "Introduction to (Social) Network Analysis in R"
author: "Dr Griffith Rees"
date: "08/04/2020"
bibliography: course.bib
output:
  html_document:
    toc: true
    number_sections: true
    # toc_depth: 2 (defaults to 3)
    toc_float: true
    includes:
      before_body: doc_prefix.html
link-citations: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

if (!require(igraph)) {
    install.packages("igraph")
}
```

# Network Data

Today we're going to demonstrate one way of loading network data, visualising it and how the network structure can help answer research questions.

## What is a Network?

A network is a way of representing how things are connected (or not). They can be social (who tweets to who), economic (which company employs which people), engineering (which parts were used in each product) etc. The representing of these connetions is a network--also call a [graph](https://en.wikipedia.org/wiki/Graph_(discrete_mathematics)) in mathematics, (confusing eh?)--and it's a tool, often applied to help quantiatively (and sometimes qualitatively) answer research questions. 

## Installing R and iGraph

We're going to focus on the iGraph R package for today. There are [*many* other options](#na-packages): 

`igraph` is a fairly comprehensive package for getting started. I recommend looking through the `igraph` documentation for your particular interests before trying the other packages, and feel free to email me [griffith.rees@sheffield.ac.uk](mailto:griffith.rees@sheffield.ac.uk)

<!--Survey: who's installed R-->

Let's just make sure that works for everyone. For those that have got it installed and running feel free to continue further down this handout to play around with the data.

Please go to <https://rstudio.com/products/rstudio/download/#download>

It should fit the operating system your using (Windows, Linux or Mac) and download `RStudio`. If you could fill in the survey when that's done that'd be great

Now let's install <https://igraph.org/r/>

```{r, eval=FALSE}
install.packages("igraph")
```

and then load it in your `R` console.

```{r, include=FALSE}
library(igraph)
```

<!--Survey of people having completed this.-->


# Networks (aka *graphs*)

In this section we will be using a small network that indicates [interactions in the movie Star Wars Episode IV](http://evelinag.com/blog/2015/12-15-star-wars-social-network/). Available here: <https://github.com/pablobarbera/data-science-workshop>. Here, each node is a character and each edge indicates whether they appeared together in a scene of the movie. Edges here are thus _undirected_ and they also have weights attached, since they can appear in multiple scenes together.

The first step is to read the list of edges and nodes in this network:

```{r}
edges <- read.csv("data/star-wars-network-edges.csv")
head(edges)
nodes <- read.csv("data/star-wars-network-nodes.csv")
head(nodes)
```
For example, we learn that C-3PO and R2-D2 appeared in 17 scenes together.

How do we convert these two datasets into a network object in R? There are multiple packages to work with networks, but the most popular is `igraph` because it's very flexible and easy to do, and in my experience it's much faster and scales well to very large networks. Other packages that you may want to explore are `sna` and `networks`. We'll see an example later in this class, because some network modeling packages are not compatible with `igraph`.

Now, how do we create the igraph object? We can use the `graph_from_data_frame` function, which takes two arguments: `d`, the data frame with the edge list in the first two columns; and `vertices`, a data frame with node data with the node label in the first column. (Note that igraph calls the nodes `vertices`, but it's exactly the same thing.)

```{r, message=FALSE}
g <- graph_from_data_frame(d=edges, vertices=nodes, directed=FALSE)
g
```

What does it mean?

* `U` means undirected  
* `N` means named graph  
* `W` means weighted graph  
* `22` is the number of nodes  
* `60` is the number of edges  
* `name (v/c)` means _name_ is a node attribute and it's a character  
* `weight (e/n)` means _weight_ is an edge attribute and it's numeric  

This is how you access specific elements within the igraph object:

```{r}
V(g) # nodes
V(g)$name # names of each node
vertex_attr(g) # all attributes of the nodes
E(g) # edges
E(g)$weight # weights for each edge
edge_attr(g) # all attributes of the edges
g[] # adjacency matrix
g[1,] # first row of adjacency matrix
```


## Basic undirected network
```{r undirected, echo=FALSE, message=FALSE, warning=FALSE, }
g1 <- graph( edges=c("A","B", "B","C", "C", "A"), directed=FALSE ) 
plot(g1, vertex.size=100, vertex.label.cex=5, edge.width=5)
```

- A, B and C are *nodes* or *vertices* (people, companies, countries etc.)
- Connections are often called *ties* or *edges* (friendship, communication, contracts, kinship etc.)
- Undirected
$$A \ue B;B \ue C; C \ue A$$

## Advantages

* A way of representing a system (social, economic, digital, biological etc.)
* Can help explore and model
    + Properties [@wasserman_social_1994]
    + Visualisations [@kerren_multivariate_2014]
    + Dynamics [@saavedra_asymmetric_2008]
    + Simulations [@bearman_chains_2004]
    
## Risks

* Complexity and size of data
* Sampling issues: **not law of large numbers** [@browne_snowball_2005]
* Pleathora of options [@butts_revisiting_2009]
* Causal claims [@vanderweele_social_2013]
    
# Nodes

## Category, Type & Attribute
```{r types, echo=FALSE, message=FALSE, warning=FALSE, }
V(g1)[0:1]$color <- "green"
plot(g1, vertex.shape=c("circle", "square", "circle"), vertex.size=c(100, 100, 160), vertex.label.cex=5, edge.width=5)
```

- Category or group (club membership, kinship etc.)
- Type (employee, firm, region etc.)
- Attribute (age, ethnicity, size etc.)

## Structure of Romantic and Sexual Relations

<!--Could add bipartite here-->

# Edges

## Direction, Weight & Type
```{r directed, echo=FALSE, message=FALSE, warning=FALSE, }
edge.scale <- c(7, 28, 3, 3, 5) 
arrow.scale <- edge.scale/1.2
g1 <- graph( edges=c("A","B", "B","C", "C", "A", "A", "C", "B", "B"), 
             directed=TRUE ) 
E(g1)$color <- c("blue", "red", "blue", "blue", "blue")
plot(g1, vertex.size=100,
     vertex.label.cex=5,
     edge.width=edge.scale,
     edge.arrow.size=arrow.scale,
     )
```

- Direction (transaction, contract, message sender vs reciever etc.)
- Weight (amount of transaction, length of contract, frequency of message etc.)
- Type (type of transaction, cateory, texts vs tweets etc.)
- $A \de B; B \de C; C \leftrightarrow A; B \de B$

## Too impressive to be true?

<iframe width="560" height="315" src="http://www.youtube.com/embed/8aEtyRD1j5U?rel=0" frameborder="0" allowfullscreen></iframe>

## That's what lots have said[^1]...

- *Homophily and Contagion Are Generically Confounded in Observational Social Network Studies* [@shalizi_homophily_2011]
- *The “unfriending” problem: The consequences of homophily in friendship retention for causal estimates of social influence* [@noel_unfriending_2011]
- *Is obesity contagious? Social networks vs. environmental factors in the obesity epidemic* [@cohen-cole_is_2008]
- *The spread of obesity in a large social network over 32 years* [@mercken_dynamics_2010][^2]
- ... and a reply: Christakis & Fowler [-@christakis_social_2013]

[^1]: See <https://statmodeling.stat.columbia.edu/2011/06/10/christakis-fowl/> for some summary
[^2]: Full disclosure: one of my examiners was a co-author on one of these papers...

# Types of Nodes 

Two types of Nodes (aka Bimodal or Bipartite Networks)

So far we've looked at nodes which vary in attributes (like name) but not in terms of types. A lot of work has been done on networks with 

## How to enforce different types of nodes


# What else?

## There's lots

- Negative ties (few studies but a recent [*Social Networks* special issue](https://www.sciencedirect.com/journal/social-networks/vol/60))
- [Erdős–Bacon number](https://en.wikipedia.org/wiki/Erd%C5%91s%E2%80%93Bacon_number)

## Useful links

* Most of this comes from/is inspired by Katherine Ognyanova's
  + [**Brilliant** visualisation guide](http://www.kateto.net/wp-content/uploads/2019/06/Sunbelt%202019%20R%20Network%20Visualization%20Workshop.pdf) 
  + Excellent github repo <https://github.com/kateto/R-Network-Visualization-Workshop>
* A bunch of useful links to papers/software etc. <https://github.com/briatte/awesome-network-analysis>
  
# Thanks

<!--## Another network time series...
![Network dynamics of creativity](collab_canvas.mov){ width=60% } -->
# Appendices

## Network Analysis Packages {#na-packages}

* R
    + [iGraph](https://igraph.org/r/) built on a C++ library, easier to get started and pre-dates ggplot so provides it's own visualisation options
    + [statnet](http://statnet.org/) focused particularly on [exponential random graphs (ERG) models](https://en.wikipedia.org/wiki/Exponential_random_graph_models), maintained by [Prof Carter T. Butts](https://scholar.google.co.uk/citations?user=-VGAs1cAAAAJ&hl=en)
    + [ggraph](https://ggraph.data-imaginist.com/) network visualisation in a way similar to [ggplot](https://ggplot2.tidyverse.org/reference/ggplot.html), maintained by [Thomas D Pedersen](https://www.data-imaginist.com/)
    + [Rsiena](https://www.stats.ox.ac.uk/~snijders/siena/) Actor-oriented logitudinal models (ERG), created by [Tom Snijders](https://cran.r-project.org/web/packages/RSiena/index.html)
* Python (my language of choice)
    + [networkx](https://networkx.github.io/) very old, not very fast but very flexible (my package of choice for ideas)
    + [igraph](https://igraph.org/python/) Effectively the same as the R package (both use the same underlying C++ library)
    + [graph-tool](https://graph-tool.skewed.de/) Much newer, very speed efficient (also C++ underneath)
* Windows
    + [UCINET](https://sites.google.com/site/ucinetsoftware/home) windows, developed and maintaied by (people including head of group at Manchester)

## References
